\documentclass{article}
\usepackage{graphicx} % Required for inserting images
\usepackage{mathtools}
\usepackage{mathrsfs}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{booktabs}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[ruled, vlined, linesnumbered]{algorithm2e}
\usepackage[margin=2cm]{geometry}
 
\SetKwInput{KwInput}{Input}
\SetKwInput{KwOutput}{Output}
\SetKwComment{Comment}{$\triangleright$\ }{}
 
\title{Regime Detection using Sliced Wasserstein k-means clustering with Monte Carlo}
\author{Oscar Peyron \\ \small \textit{ADIA Lab - University College London}}
\date{February 2026}

\begin{document}
\maketitle
\section{Abstract}  
This thesis investigates the topic of Regime Detection using an interpretable Machine Learning algorithm. 
In particular, the focus of this thesis is to produce a tractable algorithm for detecting regimes in financial markets based on multi-dimensional time series.
This thesis, provides an enhanced algorithm to cluster the multi-dimensional returns distributions. Depending, on the the characteristics of the distributions, the algorithm
can infer which regime the market is in without pre-specified assumptions on the returns distribution.
Specifically, it uses the Wasserstein distance as a metric to compare different regimes. 
In higher-dimensional settings, one can extend this concept using the Sliced Wasserstein distance and use an algorithm based on this metric to cluster different market regimes (Sliced Wasserstein k-means clustering algorithm).
The following thesis is structured as follows believed to be the most explanatory and logical way to present the research.



\section{Theoretical Background}

This chapter will focus on the theoretical framework to understand the Wasserstein distance and its extension to the Sliced Wasserstein distance.
It will also provide the mathematics behind the implied probability derived from the Wasserstein distance to produce predictive signals for trading strategies. 
Finally, we will provide some background on the statistical tests we used to compare the performance of the different trading strategies (i.e Ledoit-Wolf test on Sharpe ratios). 

\subsection{Theoretical Framework for Wasserstein Distance and Wasserstein Barycenters}
In this chapter we consider $\mathcal{X} = \mathbb{R}^d$ with $d$ the dimension of the returns vector. 
\begin{paragraph}{Definition: Data Streams}
A set $\mathcal{S}$ of streams of data over $\mathcal{X}$ is defined as: $$S(\mathcal{X}) = \left\{(X_1, X_2, \dots, X_N) | X_i \in \mathcal{X}, N \in \mathbb{N} \right\}.$$ 
For instance in this research we consider the following: 
\begin{equation}
    S = (S_1, S_2, \dots, S_N) \in \mathcal{S}(\mathbb{R}^d)
\end{equation}
where $S_i$ might be the the price vector of the different assets at time $i$.
\cite{HorvathEtAl2024}
\end{paragraph}

\begin{paragraph}{Definition: Lifting Transformation}
A lifting transformation $l$ is a map that maps the set of streams $\mathcal{S}(\mathcal{X})$ into the set of streams of streams: 
\begin{equation}
l: \left\{
\begin{aligned}
\mathcal{S}(\mathcal{X}) \rightarrow \mathcal{S}(\mathcal{S}(\mathcal{X})) \\ 
S \mapsto l(S) = (l_1, l_2, \dots, l_M)
\end{aligned}
\right.
\end{equation}
where $M>1$.  \cite{HorvathEtAl2024}
\end{paragraph}

The example from \cite{BonnierKidger2019} is the sliding window transformation: 
\begin{equation}
    l^m(\mathbf{x}) = (x_{1+h_2 (m-1)}, x_{2+h_2 (m-1)}, \dots, x_{h_1+h_2 (m-1)}) \text{ for } m = 1, 2, \dots, M
\end{equation}
with $M := \left\lfloor\frac{N-h_1-h_2}{h_2}\right\rfloor$ and $h_1, h_2 \in \mathbb{N}$ the window size and the step size respectively.

\paragraph{Definition: Discrete uniform distribution}
Given a stream generated by the lifting process $l(S) = (l^m(\mathbf{x})_i, : i = 1, 2, \dots, h_1)$, we can define the discrete uniform distribution $\mu^S$ as follows:
\begin{equation}
\mu_m = \frac{1}{h_1} \sum_{i=1}^{h_1} \delta_{l^m(\mathbf{x})_i}
\end{equation}
where we have that $\delta_x$ is the Dirac delta at point $x$. \cite{LuanHamp2025}
This definition allows us to introduce the set of empirical distributions $\mathcal{K} = \{\mu_j\}_{1 \le j \le M}$ where $\mu_j \in \mathcal{P}(\mathbb{R}^d)$ with $\mathcal{P}(\mathbb{R}^d)$ the set of probability measures on $\mathbb{R}^d$ with finite $p$-th moment. \cite{LuanHamp2025}. 


\paragraph{Definition: Wasserstein distance $\mathcal{W}_p$}
Let $\mu$ and $\nu$ be two probability measures on $\mathbb{R}^d$ with finite $p$-th moment defined as the following infimum over the joint distributions $(X,Y)$ of $d$-dimensional random vectors $X$ and $Y$. The Wasserstein distance of order $p$ between $\mu$ and $\nu$ is defined as:
\begin{equation}
\mathcal{W}_p(\mu, \nu) =  \inf_{(X,Y) \sim \Gamma(\mu, \nu)} \left( \mathbb{E}\left[ \|X - Y\|^p \right] \right)^{1/p}
\end{equation}
where $\Gamma(\mu, \nu)$ is the set of all couplings of $\mu$ and $\nu$, i.e., the set of all probability measures on $\mathbb{R}^d \times \mathbb{R}^d$ with marginals $\mu$ and $\nu$.
\cite{LuanHamp2025}

\paragraph{Definition: Wasserstein barycentre $\bar{\mu}$}
Given a family probability measures $\mathcal{K}$ in $\mathbb{R}^d$ with finite $p$-th moment, the Wasserstein barycentre $\bar{\mu}$ is defined as the probability measure that minimizes the sum of squared Wasserstein distances to the given measures:
\begin{equation}
\bar{\mu} = \arg\min_{\nu \in \mathcal{P}(\mathbb{R}^d)} \sum_{\mu_m \in \mathcal{K}} \mathcal{W}_p(\nu, \mu_m)
\end{equation}
where $\mathcal{P}(\mathbb{R}^d)$ is the set of probability measures on $\mathbb{R}^d$ with finite $p$-th moment. \cite{LuanHamp2025}

\paragraph{Projected Empirical Measure}
Let $\theta \in \mathbb{S}^{d-1}$  with $\mathbb{S}^{d-1}$ the unit sphere in $d$ dimensions.  
The projected empirical measure $\mu'(\theta) \in \mathcal{P}(\mathbb{R})$ is defined as:
\begin{equation}
\mu'(\theta) = \frac{1}{N} \sum_{i} \delta_{\langle x_i, \theta \rangle}
\end{equation}
with $x_i$ the atoms of the original empirical measure $\mu \in \mathcal{P}(\mathbb{R}^d)$ and $\langle \cdot, \cdot \rangle$ the standard inner product in $\mathbb{R}^d$. \cite{LuanHamp2025}


\paragraph{Definition: The sliced Wasserstein distance $\mathcal{\bar{W}}_p$}
Given two distributions $\mu, \nu \in \mathcal{P}(\mathbb{R}^d)$ with $d>1$, the sliced Wasserstein distance is defined as follows: 
\begin{equation}
\mathcal{\bar{W}}_p(\mu, \nu) = \int_{\mathbb{S}^{d-1}} \mathcal{W}_p(\mu'(\theta), \nu'(\theta)) d\theta
\end{equation}
with $\mathcal{W}_p(\mu'(\theta), \nu'(\theta))$ the Wasserstein distance in d=1 between the projected empirical measures $\mu'(\theta)$ and $\nu'(\theta)$ on the direction $\theta$. \cite{LuanHamp2025}


\paragraph{Definition: The sliced Wasserstein Barycenter}

%%% explain more in detail was is the Wasserstein distance and the basics 


\subsection{Theoretical Framework for Sliced Wasserstein (SW) with UnifOrtho Projections}
This section will provide insights on the practical implementation of the Sliced Wassestein k-mean clusterings algorithms and limitations to the 
its practical implementations in the literature. Notably, the low generalizable nature of using pre-defined projection vectors ${\theta^l}$  for the Sliced Wasserstein can be solved 
by generating them using the UnifOrtho method \cite{PetrovicBardenetDesolneux2025}. 
It provides the following advantages compared to standard Monte Carlo methods. 
\begin{itemize}
    \item \textbf{Variance reduction via Negative Negative Dependance}
    \item \textbf{Superior performance in high dimensions} : The UnifOrtho is therefore perfectly adapted to working with multi-dimensional time series for the SW k-means clustering algorithm.
    \item \textbf{Clustering Stability}
\end{itemize}

Some useful theoretical results on variance reduction and computational cost on the UnifOrtho method will therefore be provided to support the choice of this method for the future sections.

The following pseudocode is based on the work by \cite{PetrovicBardenetDesolneux2025} on Repulsive Monte Carlo Methods and \cite{LuanHamp2025} on Sliced Wasserstein k-means clustering for regime detection. 

\begin{algorithm}[H]
\caption{sWk-means clustering with UnifOrtho Projections}\label{alg:swkmeans}

\KwInput{Data stream $S \in \mathcal{S}(\mathbb{R}^d)$,\; number of clusters $K$,\; number of projections $L$,\; tolerance $\epsilon$ for convergence}
\KwOutput{$K$ cluster assignments and corresponding 1d barycenters}

\BlankLine

\Comment{Preprocessing}
Calculate $l(r^S)$ given $S \in \mathcal{S}(\mathbb{R}^d)$ and define family of empirical distributions $\mathcal{K} = \{\mu_j\}_{1 \le j \le M}$\;

\BlankLine
\Comment{Step 1: UnifOrtho Projection Generation}
$k \leftarrow \lceil L/d \rceil$ \Comment*[r]{number of orthogonal bases needed}
$\Theta \leftarrow \emptyset$\;

\For{$i = 1$ \KwTo $k$}{
    Sample a random matrix $Z_i \in \mathbb{R}^{d \times d}$ with i.i.d.\ $\mathcal{N}(0,1)$ entries\;
    Compute the QR decomposition: $Z_i = Q_i R_i$\;
    Let $\Lambda_i$ be a diagonal matrix where $\Lambda_{i, jj} = \text{sgn}(R_{i, jj})$\;
    $U_i \leftarrow Q_i \Lambda_i$ \Comment*[r]{$U_i$ is Haar-distributed on $O(d)$}
    Add the $d$ column vectors of $U_i$ to $\Theta$\;
}
$\{\theta^1, \dots, \theta^L\} \leftarrow \Theta[1:L]$ \Comment*[r]{select exactly $L$ directions}

\BlankLine
\Comment{Step 2: Projection \& K-Means Iteration}
Calculate projected distributions $\{\mu_j'(\theta^l)\}_{1 \le j \le M}$ for all $l=1 \dots L$\;
Initialize cluster assignments $C_1, \dots, C_K$\;

\While{convergence criterion $> \epsilon$}{
    \For{$k = 1$ \KwTo $K$}{
        Compute the 1d Wasserstein barycenter $\bar{\mu}_k(\theta^l)$ for each projection $l$ using empirical distributions assigned to $C_k$\;
    }
    \For{$j = 1$ \KwTo $M$}{
        Assign $\mu_j$ to the cluster $C_{k^*}$ where $k^* = \arg\min_{k} \sum_{l=1}^L W_p^p\!\left(\mu_j'(\theta^l),\, \bar{\mu}_k(\theta^l)\right)$\;
    }
}

\Return{$C_1, \dots, C_K$ and barycenters $\{\bar{\mu}_k(\theta^l)\}$}
\end{algorithm}

\subsection{Theoretical Framework for Implied Probability Regime Detection Strategy}

\begin{paragraph}{Theoretical Background}
We use implied probability measure to compute wether on thinks the market is going to become bear or bull. 
\paragraph{Likelihood Function}
It uses SW-distance to every centroid to every centroid and then convert the likelihood via a softmax over the negative distances. 
We introduce a temperature parameter $\tau$ 
$$
\mathcal{L}(\text{regime } k \mid x_m) = \frac{e^{-d(x_m, c_k) / \tau}}{\sum_{j} e^{-d(x_m, c_j) / \tau}}
$$

\paragraph{Empirical Transition Matrix Markov Prior}
A market in regime i is highly likely to stay in regime i tomorrow. We model this as a first-order Markov Chain using the historical labels. 
Probability of switching from i to j is: 
$$
\pi_{i,j} = P(Z_m = j \mid Z_{m-1} = i) = \frac{\text{Count}(i \to j)}{\sum_{x=1}^{K} \text{Count}(i \to x)}
$$
If the market was $z_{current}$ our prior belief that state M is going to be at regime $k$ is : 
$$
\mathbb{P}_{prior}(k) = \pi_{z_{current},k}
$$

Then we simply use Bayes rule to derive the following :
$$
\mathbb{P}_{Bayes}(k) = \frac{\mathbb{P}_{prior}(k) \cdot \mathcal{L}(k \mid X_m)}{\sum_{j=1}^{K} \mathbb{P}_{prior}(j) \cdot \mathcal{L}(j \mid X_m)}
$$

\paragraph{Switch Probability}
Finally the switch probability does the following : 
$$
\mathbb{P}_{switch} = 1- \mathbb{P}_{Bayes}(z_{current})
$$

\paragraph{Use Gradient} 
One may also want to take into account the fact that one is moving away or getting closer to a certain regime. 

We consider a lookback window of size $\omega$, and one computes the linear trend slope of the Wasserstein distance to each centroid $k$. 
Let $t \in \{1,2, \dots, \omega \}$
We get using  OLS, $\beta_{k}$:

$$
\beta_k = \frac{\text{Covariance}(t, \, \mathcal{W}_2(X_{M-w+t}, \, C_k))}{\text{Variance}(t)}
$$

\begin{itemize}
    \item if $\beta_{k} < 0$ the distance to regime k is shrinking (we are getting close to regime k)
    \item if $\beta_{k} > 0$ the distance to regime k is growing  (we are drifting away from regime k)
\end{itemize}

We apply softmax to get a probability: 
$$
P_{Grad}(k) = \frac{\exp\left(-\frac{\beta_k}{\tau}\right)}{\sum_{j=1}^{K} \exp\left(-\frac{\beta_j}{\tau}\right)}
$$

The final switch probability using gradient is therefore: 

$$
\mathbb{P}_{switch} = 1- \left[(1-\alpha)\mathbb{P}_{Bayes}(z_{current}) + \alpha \mathbb{P}_{Grad}(z_{current}) \right]
$$

with $\alpha \in [0:1]$ a weight. 

\end{paragraph}

\subsection{Theoretical Framework for the Ledoit-Wolf Sharpe ratio test}

% To be added
Need to read the paper \cite{LedoitWolf2008}. 



\section{Design methodology for Monte Carlo for SW - Synthetic Data Testing}
The focus of this chapter will be on testing the algorithm on syntethic data to understand the behaviour of the algorithm and its performance in different settings.
The algorithm will be tested on 3 dimensional data followed by 4, 5 dimensional.  Performance between original and UniOrtho method will be compared and tests will span from data with multiple correlation factors 
and with diffrent target number of clusters. Finally, in terms of time-horizon, the study will focus on the equivalent of  20 years synthetic data with 7 observations a day (252 trading days a year) to be as close as possible to real world data.


Different performance metrics will be used for the regime detection : 
\begin{itemize}
    \item Total Accuracy (TA)
    \item Mean centroid-centroid distance (MCCD)
    \item Point-to-centroid distance (PCD) 
\end{itemize}

The synthetic data will be generated similarly to previous papers \cite{HorvathEtAl2024} and \cite{LuanHamp2025}
using Geometric Brownian Motion (GBM) paths $S_t/S_{t-1} = \exp(r^S_t)$ with $S_t = (S_t^1, \dots, S_t^d)$ and $r^S_t = (r^{S^{(1)}}_{t}, \dots, r^{S^{(d)}}_{t})$ 
where : 
\begin{equation}
    r^{S^{(i)}}_{t} \sim \mathcal{N}\left((\mu-\sigma^2/2)\bm{1}dt, \bm{\Sigma} dt \right)
\end{equation}.
with 
\begin{equation}
\Theta = (\mu, \sigma^2)
\end{equation}

And the covariance matrix $\bm{\Sigma}$ is defined as follows :
\begin{equation}
\Sigma_{ij} = \sigma^2 \left( \delta_{ij} + (1- \delta_{ij})\rho \right)
\end{equation}
The $\delta_{ij}$ is the Kronecker delta, $\sigma^2$ is the variance of the returns and $\rho$ is the correlation between the different assets.

Assuming, we want to cluster two different regimes (K=2), we define, as in previous papers \cite{LuanHamp2025}, Bull and Bear regimes with the following parametrization : 
\begin{itemize}
    \item $\Theta_{\text{bull}} = (0.02, 0.2)$
    \item $\Theta_{\text{bear}} = (-0.02, 0.3)$
\end{itemize}
For K = 3, similarly to \cite{LuanHamp2025} we will consider the \textit{moon-shape structure} generated using \texttt{datasets.make\_moons} function from the \texttt{sklearn} library.

This is a summary similary to the one provided in \cite{LuanHamp2025} for the different market settings on which we are going to test the sWk-means clustering algorithm with UnifOrtho projections :

\begin{table}[h]
    \centering
    \caption{Summary of synthetic dataset parameters with $d\in \left\{3,4,5\right\}$}
    \label{tab:dataset_parameters}
    \begin{tabular}{lccc}
        \toprule
        Type & Regime Bull & Regime Bear & Regime III (Bear/Moon)  \\ 
        \midrule
        Type A (K=2) & $\Theta_{\text{bull}}; \rho = +1/2$ & $\Theta_{\text{bear}}; \rho = +1/2$ & N/A \\
        Type B (K=2) & $\Theta_{\text{bull}}; \rho = +1/2$ & $\Theta_{\text{bull}}; \rho = -1/2$ & N/A \\
        Type C (K=3) & $\Theta_{\text{bull}}; \rho = +1/2$ & $\Theta_{\text{bear}}; \rho = +1/2$ & $\Theta_{\text{bear}}; \rho = -1/2$ \\
        Type D (K=3) & $\Theta_{\text{bull}}; \rho = +1/2$ & $\Theta_{\text{bear}}; \rho = +1/2$ & $\Theta_{\text{moon}}; \rho = +1/2$ \\
        \bottomrule
    \end{tabular}
\end{table}

\section{Application of sliced-SW k-means clustering with UnifOrtho projections on real world data sets}

Finally, the chapter will focus on the application of the novel algorithm on real world data sets for different types of financial markets. 
As above, the time span, wil focus on 15 to 20 years of data. The target data sets include the following : 

\begin{itemize}
    \item FX Spot Series Data (4D)  
        \subitem EURUSD, AUDUSD, USDJPY, USDCHF, USDCAD (Major Currency Cluster) (15 min data)
    \item Equity Series Data (3D)
        \subitem S\&P 500, NASDAQ, Dow Jones (US Equity Cluster) (daily data)
        \subitem CAC 40, DAX40, FTSE 100 (European Equity Cluster) (daily data)
        \subitem Nikkei 225, Hang Seng, Kospi 200 (Asian Equity Cluster) (daily data)

    \item  Commodity Series Data (4D and 5D)
        \subitem WTI Crude Oil, Brent Crude Oil, Natural Gas, Heating Oil. 
        \subitem Copper Aluminum, Zinc, Nickel, Lead. 
\end{itemize}

\subsection{Performance analysis of regime detection using trading strategies}
\begin{itemize}
\item Long/Short strategy based on majority regime detection signals
\item Long/Short strategy based on Implied probability signals (using the distance to the centroids as a proxy for the confidence of the regime detection)
\item Ensemble Methods: combining the above two strategies using a weighted average of the signals or an adaptative weighting scheme to generate the final trading signal.
\end{itemize}

Tested trading strategies on FX and Equity data sets. Performance is compared to a long only and short-only approach. 
Moreover, for equities, we look at the performance for different clusters (US, Europe, Asia) to see if the performance of the regime detection is better in some markets than others.
We also test, trading algorithm that use american regime detection to trade in Asian and European markets to see if the regime detection is more generalizable in some markets than others.

\subsubsection{Standard SW k-means trading strategy}
Pseudo code for this strategy is provided here : 


\begin{algorithm}[H]
\caption{\textsc{Long\_Strat\_Unifortho} --- Binary Regime Strategy}\label{alg:unifortho}
 
\KwInput{Initial capital $C_0$,\; number of simulations $N_S$,\; price matrix $S\in\mathbb{R}^{T\times d}$,\; trimming parameters $h_1, h_2$,\; window size $w$,\; number of clusters $K$,\; risk metric $m$,\; majority look-back $\ell$}
\KwOutput{Portfolio value path $V$}
 
\BlankLine
 
$R \leftarrow \text{PctReturns}(S)$ \Comment*[r]{$R_t = S_t / S_{t-1} - 1$}
$\theta \leftarrow \mathbf{1}_d / \sigma_d$ \Comment*[r]{inverse vol weighted returns}
$r^{(\text{pf})} \leftarrow R\,\theta$ \Comment*[r]{portfolio return series}
$V \leftarrow [C_0]$\;
$n \leftarrow \lfloor T / w \rfloor$\;
 
\BlankLine
\For{$i = 0$ \KwTo $n - 2$}{
    $s \leftarrow i \cdot w$, \quad $e \leftarrow (i+1)\cdot w$\;
    $W \leftarrow S[s : e]$ \Comment*[r]{current analysis window}
 
    \BlankLine
    \Comment{Step 1 --- Wasserstein-based regime detection}
    $\hat{F},\; \boldsymbol{\mu},\; \boldsymbol{\lambda} \leftarrow \textsc{MaxMCCD\_UnifOrtho}(N_S,\, W,\, K,\, L,\, \varepsilon,\, h_1,\, h_2,\, m)$\;
 
    \BlankLine
    \Comment{Step 2 --- Majority-vote regime classification}
    \eIf{$\ell > |\boldsymbol{\lambda}|$}{
        $\hat{k} \leftarrow \arg\max_k \;\text{count}(\boldsymbol{\lambda} = k)$\;
    }{
        $\hat{k} \leftarrow \arg\max_k \;\text{count}(\boldsymbol{\lambda}_{[-\ell:]} = k)$\;
    }
 
    \BlankLine
    \Comment{Step 3 --- Generate binary signal}
    \eIf{$\hat{k} = 1$}{
        $\sigma \leftarrow +1$ \Comment*[r]{Bullish $\Rightarrow$ Long}
    }{
        $\sigma \leftarrow -1$ \Comment*[r]{Bearish $\Rightarrow$ Short}
    }
 
    \BlankLine
    \Comment{Step 4 --- Compound portfolio over next window}
    \For{each $r_t \in r^{(\mathrm{pf})}[e : e + w]$}{
        $V.\text{append}\!\Big(V[-1]\;\times\;\big(1 + \sigma\, r_t\big)\Big)$\;
    }
}
 
\Return{$V$}
\end{algorithm}
 
\vspace{1.5cm}

\subsubsection{Implied Probability SW k-means trading strategy}

\begin{algorithm}[H]
\caption{\textsc{Long\_Strat\_Implied} --- Implied-Probability Regime Strategy}\label{alg:implied}
 
\KwInput{Initial capital $C_0$,\; simulations $N_S$,\; price matrix $S\in\mathbb{R}^{T\times d}$,\; trimming $h_1, h_2$,\; window $w$,\; clusters $K$,\; metric $m$,\; signal type $\tau \in \{\texttt{continuous},\, \texttt{hysteresis},\, \texttt{conviction}\}$,\; entry threshold $\alpha_e$,\; hold threshold $\alpha_h$,\; look-back $\ell$,\; gradient flag and weight $(\texttt{g},\, \omega_g)$}
\KwOutput{Portfolio value path $V$,\; signal history $\Sigma$}
 
\BlankLine
 
$R \leftarrow \text{PctReturns}(S)$, \quad $\theta \leftarrow \mathbf{1}_d \cdot 1/\sigma_d$
%inverse vol weighted returns
\quad $r^{(\text{pf})} \leftarrow R\,\theta$\;
$V \leftarrow [C_0]$, \quad $\Sigma \leftarrow [\,]$, \quad $n \leftarrow \lfloor T/w \rfloor$\;
\BlankLine
\For{$i = 0$ \KwTo $n - 2$}{
    $s \leftarrow i\cdot w$, \quad $e \leftarrow (i+1)\cdot w$, \quad $W \leftarrow S[s:e]$\;
 
    \BlankLine
    \Comment{Step 1 --- Regime detection (same as Algorithm~\ref{alg:unifortho})}
    $\hat{F},\;\boldsymbol{\mu},\;\boldsymbol{\lambda} \leftarrow \textsc{MaxMCCD\_UnifOrtho}(N_S,\,W,\,K,\,L,\,\varepsilon,\,h_1,\,h_2,\,m)$\;
 
    \BlankLine
    \Comment{Step 2 --- Implied transition probabilities}
    $\mathbf{P},\; p_{\text{switch}},\; \mathbf{T},\; \boldsymbol{\pi} \leftarrow \textsc{ComputeImpliedProba}(\hat{F},\;\boldsymbol{\mu},\;\boldsymbol{\lambda},\;\ell,\;\texttt{g},\;\omega_g)$\;
    $\hat{k} \leftarrow \arg\max_k\;\text{count}(\boldsymbol{\lambda}_{[-h_2:]} = k)$\;
 
    \BlankLine
    \Comment{Step 3 --- Signal generation (depends on $\tau$)}
    \uIf{$\tau = \texttt{continuous}$}{
        $\sigma \leftarrow \pi_{\text{bull}} - \pi_{\text{bear}}$\;
        \If{$|\sigma| < \delta$}{$\sigma \leftarrow 0$} \Comment*[r]{dead zone}
    }
    \uElseIf{$\tau = \texttt{hysteresis}$}{
        $\sigma_{\text{prev}} \leftarrow \Sigma[-1]$ \textbf{or} $0$\;
        \uIf{$\hat{k} = \text{Bull}$}{
            \lIf{$p_{\text{switch}} \geq \alpha_e$ \textbf{and} $\sigma_{\text{prev}} \geq 0$}{$\sigma \leftarrow -1$}
            \lElseIf{$p_{\text{switch}} < \alpha_h$ \textbf{and} $\sigma_{\text{prev}} < 0$}{$\sigma \leftarrow +1$}
            \lElse{$\sigma \leftarrow 0$}
        }
        \Else(\Comment*[f]{$\hat{k} = \text{Bear}$}){
            \lIf{$p_{\text{switch}} \geq \alpha_e$ \textbf{and} $\sigma_{\text{prev}} \leq 0$}{$\sigma \leftarrow +1$}
            \lElseIf{$p_{\text{switch}} < \alpha_h$ \textbf{and} $\sigma_{\text{prev}} > 0$}{$\sigma \leftarrow -1$}
            \lElse{$\sigma \leftarrow 0$}
        }
    }
    \ElseIf{$\tau = \texttt{conviction}$}{
        $d_k \leftarrow \begin{cases} +1 & \hat{k} = \text{Bull} \\ -1 & \hat{k} = \text{Bear}\end{cases}$\;
        $c \leftarrow 1 - 1.5\, p_{\text{switch}}$ \Comment*[r]{conviction decays with switch risk}
        $\sigma \leftarrow d_k \cdot c$\;
    }
    $\Sigma.\text{append}(\sigma)$\;
 
    \BlankLine
    \Comment{Step 4 --- Compound portfolio over next window}
    \For{each $r_t \in r^{(\mathrm{pf})}[e : e+w]$}{
        $V.\text{append}\!\Big(V[-1]\;\times\;\big(1 + \sigma\, r_t\big)\Big)$\;
    }
}
\Return{$V,\;\Sigma$}
\end{algorithm}


\subsection{Ledoit-Wolf test on sharpe ratios}
Conducted Ledoit-Wolf test to compare the Sharpe ratios of the different trading strategies and see if the differences in performance are statistically significant.
Conducted test with a confidence level of 95\% and reported the p-values for each pairwise comparison between the trading strategies and the long-only benchmark.
Used hac method as well as the studentized circular bootstrap approach. 

\subsection{FX Data}
I have tested different trading strategies based on regime detection using the sWk-means clustering algorithm with UnifOrtho projections and the implied probability strategy derived from the SW distance to the centroids and compared it to a long-only benchmark on the
training period from 2006 to 2020 (FX data). 


\subsubsection{Equity Data}
I have tested different trading strategies based on regime detection using the sWk-means clustering algorithm with UnifOrtho projections and the implied probability strategy derived from the SW distance to the centroids and compared it to a long-only benchmark on the
training period from 2006 to 2020 (3483 observations for the American Equity Data).

\paragraph{American Indices Equity}
No significant statistically difference in sharpe ratios betweent the trading strategies and a long-only benchmark. 


\subsection{Tuning the hyperparameters for the trading strategies}
Tested multiple values for the hyperparameters of the trading strategies, such as the entry and hold
\subsubsection{Tuning the weights for ensemble methods}

Tested multiple combination of weights for the ensemble methods, which combines the binary regime long/short stratey and the implied probability strategy with different variations (hysteresis, continuous, conviction).
The weights are ordered as following :
\newline [$w_{binary}$, $w_{implied\_probability(hysteresis)}$ $w_{implied\_probability(continuous)}$, $w_{implied\_probability(conviction)}$]. 
We have tested the following weights: 
\begin{itemize}
    \item \text{[0.25, 0.25, 0.25, 0.25]},
    \item \text{[0.30, 0.10, 0.30, 0.30]},
    \item \text{[0.20, 0.30, 0.20, 0.30]},
    \item  \text{[0.30, 0.20, 0.30, 0.20]},
    \item  \text{[0.10, 0.40, 0.10, 0.40]}
\end{itemize}

\subsubsection{Tuning the gradient weight for implied probability methods}
Tested multiple values for the gradient weight $\alpha$ in the implied probability strategy with gradient.
The values tested are the following 
\begin{equation}
    \alpha \in \{0.0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1.0\}
\end{equation}

Optimal value for $\alpha$ is determined based on the highest Sharpe ratio achieved during the backtesting period.
For the American equity data we have determined that the optimal value is $\alpha = 0.5$. As most of the variants of implied porbability strategies (continus and hysteresis), have a peak performance at $\alpha = 0.5$. 

$\alpha = 0.0$ corresponds to a pure bayesian approach. It captures current cluster membership, regime persistence and more generally where the market is. 
$\alpha = 1.0$ corresponds to a pure gradient approach. It captures the direction of the market (getting closer or further from a certain regime) and therefore captures the momentum of the market.

Hence an optimal value of $\alpha = 0.5$ suggests that  regime identification and regime momentum are equally predictive. 
We could also thinkg that $\mathbb{P}_{Bayers}$ and $\mathbb{P}_{Grad}$ are not redundant and capture different information about the market. 


\subsubsection{Tuning the temperature parameter $\tau$ and $\tau_{grad}$ for the implied probability strategy}
Peformed grid search for $\tau$ taken as adaptative to the switch probability as well as for $\tau_{grad}$ taken as adaptative to the gradient values.
The optimal values for $\tau$ and $\tau_{grad}$ are determined based on the highest Sharpe ratio achieved during the backtesting period. 
For the American equity data, we have determined that the optimal value for $\tau$ is 0.01 and the optimal value for $\tau_{grad}$ is 0.01.

\section{Experiment 5: Interpretability Analysis of Sliced Wasserstein Regime Detection Algorithm}

We label the regimes using weekly equity data (S\&P 500, NASDAQ, Dow Jones) on the period 2006 to 2020. 
We further consider the following macro feature set:
\begin{itemize}
    \item VIX Index
    \item 10-year Treasury Yield
    \item 5-year Treasury Yield
    \item 30-year Treasury Yield
    \item Fed Funds Rate
    \item Manufacturing PMI
    \item Manufacturing Employment
    \item DXY Dollar Index
    \item Gold Spot Price
    \item Brent Crude Oil Spot Price
\end{itemize}

After having regrouped the data from the following macro-indicators from the appropriate time period, we fit thefeature set to the regime labels using the following models: 
\begin{itemize}
    \item OLS Linear Regression (probabilistic target (Probability of being in bear regime))
    \item Logistic Regression (binary target (Bull/Bear))
    \item Random Forest Classifier (binary target (Bull/Bear))
\end{itemize}
In particular, we are analysing the feature importance of the different macro-indicators during the bear regimes using the SHAP analysis toolbox. 
We observe that in particular during predicted bear regimes by the sliced-Wasserstein k-means clustering algorithm, the VIX index is a particularly important feature in the regime classification for all three models. 



\newpage
\section{Appendix}
\newpage
\bibliographystyle{plain}
\bibliography{main}

\end{document}